Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-27 paper.tex
full paper read, added some comments esp. on results

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG). Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL). Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution. The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources. Effective virtualization requires \emph{oversubscription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available. As a result, a single multiprocessor may host multiple workgroups simultaneously. However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.

@@

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable for a portable environment, as excessively large work tiles risk reducing occupancy on less capable hardware.

@@

Hardware-level considerations underscore the limitations of fixed-launches. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite{} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their findings in in \cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis in \cite{sorensen2018}.

Sorensen et al. classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion. HSA guarantees termination, as the lowest ID workgroup is always given forward progress. If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.

@@

\subsection{Why is reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is typically performed by the first thread.

@@

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory. Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

We now detail the \emph{Decoupled-Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained-Propagation} phase entirely.
        \end{itemize}

  \item \textbf{Initialization:}
        \begin{itemize}
          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
          \item The first thread initializes a \emph{previous reduction} variable in registers with $\varepsilon$ to hold the accumulated reductions of preceding tiles.
          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
        \end{itemize}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}. The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

@@

Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state). That, in turn, makes it impossible to represent common ordinary 32-bit floating point values without loss of precision. While some GPUs support 64 bit atomics (and it is likely that WebGPU will include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. The finer details of this distinction are discussed next.
\section{Discussion}

@@

\subsection{Conclusion}
In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary shading languages. Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this behavior.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG). Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL)\@. \john{I don't love ``compatibility with shading languages'' but then we name a single shading language. Instead, maybe something like ``Our implementation targets portability across modern GPU hardware and software platforms'', that sort of thing?} Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting a subset of the currently \emph{occupant} subgroups on a multiprocessor for execution. The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources. Effective virtualization requires \emph{oversubscription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available. As a result, a single multiprocessor may host multiple workgroups simultaneously. However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context-switch.

@@

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable from a portability perspective, as excessively large work tiles risk reducing occupancy on less capable hardware.

@@

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

Sorensen et al.\ classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion. HSA guarantees termination, as the lowest ID workgroup is always given forward progress. If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.

@@

The viability of \emph{Decoupled Fallback} relies on the same cache locality properties as \emph{Decoupled Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to false-positive fallback operations. As a result, memory bandwidth usage increases (and we expect throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate in Section~\ref{sec:Sim-Blocking}, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory. Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{We may also obtain a workgroup-wide reduce operation by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.\john{I don't get this footnote. ``May also''? What are we actually doing? What's the point of this footnote?}}

\subsection{Decoupled Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---\emph{Not Ready} $\rightarrow$ \emph{Ready} $\rightarrow$ \emph{Inclusive}---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\varepsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

We now detail the \emph{Decoupled Fallback} procedure, presenting the algorithm from the perspective of a single workgroup. The details of tile state encoding and atomic updates are discussed in the following section and may vary depending on the data type, so for simplicity, we describe the operation as being performed by the \emph{first thread} of the workgroup:
\begin{figure}
  \centering
  \includegraphics[height=\textheight]{graphics/FlowChart.pdf}
  \caption{Flow Chart Representation of Decoupled Fallback.\label{fig:decoupled_fallback}}
\end{figure}
\begin{enumerate}
  \item[(0)] \textbf{Posting Local Reduction:}
        \begin{itemize}
          \item Each workgroup posts its local reduction---the result of the workgroup-wide scan over partials---to global memory and updates its tile state to \emph{Ready}.
          \item If it is the first work tile, its local reduction \emph{is} its inclusive reduction. In this case, the workgroup updates its tile state directly to \emph{Inclusive} and bypasses the \emph{Chained Propagation} phase entirely.
        \end{itemize}

  \item \textbf{Initialization:}
        \begin{itemize}
          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained Propagation} phase.
          \item The first thread initializes a \emph{previous reduction} variable in registers with $\varepsilon$ to hold the accumulated reductions of preceding tiles.
          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
        \end{itemize}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well-studied in the context of memory models~\cite{Alglave0, MCMutants}. The key requirements here are both control synchronization (meaning the reader does not proceed until the writer sends the message) and data synchronization (meaning that the reader reads the correct value). The latter is especially challenging in relaxed memory models as are employed on GPUs.

@@

Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state). That, in turn, makes it impossible to represent ordinary 32-bit floating point values without loss of precision. While some GPUs support 64 bit atomics (and it is likely that WebGPU will someday include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that we are inducing fallbacks, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. The finer details of this distinction are discussed next.
\section{Discussion}

@@

\subsection{Conclusion}
In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary portable shading languages. Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.
